# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

Read in activity.csv from the working directory, and define the class of each column

```{r}
classes <- c(steps = "integer", date = "Date", interval = "integer")
#read in data frame with 17568 observations
activity  <- read.csv("activity.csv", colClasses = classes)
```


## What is mean total number of steps taken per day?

```{r}
library(reshape2)
activitymelt <- melt(activity,id = c("date", "interval"), na.rm = TRUE)
#sumbyday is the sum total of steps taken per day
sumbyday <- dcast(activitymelt, date ~ variable , sum)
#meanbyday is the mean number of steps taken each day
meanbyday <- dcast(activitymelt, date ~variable, mean)

```

```{r fig.width=7, fig.height=6}
hist(sumbyday$steps/1000, xlab= "Total number of steps by day in thousands", main = NULL, breaks = 15)
```

```{r}

stepstats  <- c(mean = mean(sumbyday$steps), median = median(sumbyday$steps))
stepstats
```



## What is the average daily activity pattern?
```{r}
averageoverdays <- dcast(activitymelt, interval ~ variable, mean)

with(averageoverdays, plot(interval, steps, type = "l"))

averageoverdays[averageoverdays$steps %in% max(averageoverdays$steps),]

```
The time between 8:35 and 8:40 contains the maximum average of 206 steps.

## Imputing missing values
I am going to average the number of steps from that interval of the surrounding days
There are 8 days of missing data 288 observation/day *8 = 2304
The dates are 
- 2012-10-01
- 2012-10-08
- 2012-11-01
- 2012-11-04
- 2012-11-09
- 2012-11-10
- 2012-11-14
- 2012-11-30

```{r}
dates <- is.na(activity$steps)
avgdates <-activity$date[!dates]
dates <- activity$date[dates]
dates <- unique(dates)
avgdates <- unique(avgdates)
intervals <- unique(activity$interval)

useavgdate <- activity$date %in% avgdates
useactivity <- activity[useavgdate,]
activityavg <- melt(useactivity,id = c("date", "interval"), na.rm = TRUE)
averagetouse <- dcast(activityavg, interval ~ variable, mean)

update <- activity$date %in% dates

#activity$interval %in% intervals
#activity$steps == averagetouse$steps
#for (i in seq_along(intervals)){



for (i in seq_along(dates)){
      #print(dates[i])
      count = 0
      if ((dates[i]-1) %in% activity$date){
            #print ((dates[i]-1) %in% activity$date)
            count = count -1
            #if (activity$steps[activity$date==dates[i]-1]==NA){count = -10}
      }
      if ((dates[i]+1) %in% activity$date){
           # print ((dates[i]+1) %in% activity$date)
            count = count +1
            if (NA %in% activity$steps[activity$date==dates[i]+1]){count = +20}
      }
      #print (count)
      rolling =1
      for (j in seq_along(intervals)){
      
            if (count == -1){
                  activity$steps[activity$date==dates[i]&
                                       activity$interval==intervals[j]] <- 
                        activity$steps[activity$date==dates[i]-1&
                                       activity$interval==intervals[j]]
            }
            if (count == 1){
                  activity$steps[activity$date==dates[i]&
                                       activity$interval==intervals[j]] <- 
                        activity$steps[activity$date==dates[i]+1&
                                       activity$interval==intervals[j]]
            }
            if (count == 0){
                  
                  activity$steps[activity$date==dates[i]&
                                       activity$interval==intervals[j]] <- 
                        mean(activity$steps[activity$date==dates[i]+1&
                                       activity$interval==intervals[j]],
                             activity$steps[activity$date==dates[i]-1&
                                       activity$interval==intervals[j]])
                  #rolling =rolling+1
                  #print (rolling)
            }
            if (count == 20){
                  
                  activity$steps[activity$date==dates[i]&
                                       activity$interval==intervals[j]] <- 
                        mean(activity$steps[activity$date==dates[i]+2&
                                       activity$interval==intervals[j]],
                             activity$steps[activity$date==dates[i]-1&
                                       activity$interval==intervals[j]])
                  rolling =rolling+1
                  #print (rolling)
            }
      }
}


```


## Are there differences in activity patterns between weekdays and weekends?
```{r}
library("ggplot2")
activity$weekday <- weekdays(activity$date)
activity$weekday <- as.factor(activity$weekday)
weekend <- grep("S.+", activity$weekday)
#weekend
activitymelt <- melt(activity,id = c("date", "interval", "weekday"), na.rm = TRUE)
#sumbyday is the sum total of steps taken per day
sumbyday <- dcast(activitymelt, date ~ variable , sum)


qplot(interval, steps, data=activity, facets = .~weekday, rm.na = TRUE, geom = "line")
qplot(interval, steps, data=averageoverdays, geom = c("point", "line"))
par(mfrow = c(2, 1))
with(averageoverdays, plot(interval, steps, type = "l"))

```
Any reference to a commit’s SHA-1 hash will be automatically converted into a link to that commit on GitHub.
